Replacing handwritten signatures with open electronic signature software

### 2020-05-07 paper/main.tex
Add fixes for paper feedback

@@

\abstrakt{
So zmenami v právnom statuse elektronických podpisov vo svete sa vytvára príležitosť rozšíriť elektronické podpisovanie ako náhradu pre vlastnoručné podpisy mimo obvyklých oblastí.
Elektronické podpisy sú dôležité čoraz viac v dobe elektronickej komunikácie kedy sú vlastnoručné podpisy nepraktické z dôvodu nutnosti tlače, osobného kontaktu alebo zasielania podpísaných dokumentov.
V práci sa venujeme právnym predpokladom relevantným pre EÚ, technologickým predpokladom, prehľadu v súčasnosti dostupného softvéru a návrhu vlastnej softvérovej platformy.
Výsledkom je otvorený, používateľsky prívetivý softvér pre platformy Windows, Mac OS X a Linux.
Pomocou aplikácie je možné, v súlade s nariadením eIDAS, podpísať a overiť podpis ľubovoľného dokumentu, ktorý má rovnaký právny účinok ako osvedčený podpis.
Naše riešenie podporuje občianske preukazy Slovenskej Republiky s čipom eID bez akýchkoľvek úprav alebo nastavení, čím sa umožní ich jednoduché využitie mimo obvyklého použitia vo webových aplikáciách štátu.
Aplikácia nie je viazaná iba na slovenské eID, ale je pripravená na ďalšie použitie s rôznymi zariadeniami v rôznych štátoch a ako platforma pre iné druhy podpisov.
Súčasťou je aj webová stránka, ktorá okrem možnosti stiahnutia softvéru a stručnej dokumentácie poskytuje všeobecné informácie o elektronických podpisoch aby bol projekt čo najviac prístupný verejnosti.
Dôraz kladieme na jednoduchú rozšíriteľnosť programu vo forme jazykových prekladov, dokumentácie, funkcionality a všeobecných informácií.
}

@@

While, specifically in the European Union, they seem to be relatively extensively used in communication with the government, they do not seem to be widely adopted in an exchange between individuals or smaller organizations.
The problem is handwritten signatures are often unpractical for electronic communication due to the need to print, personally exchange, or send signed documents.

Our goal is to explore an accessible way to use electronic signatures.
In the beginning, we go over the preconditions that make it a viable alternative to the handwritten signatures.
We cover the legal preconditions relevant for the European Union and Slovakia in particular.
Also, we explain technical preconditions focusing on the cryptographic concepts like asymmetric cryptography, timestamping, hashing, and standards that comply with the law within the European Union.
Secondly, we do a review of the existing software.
We go quickly over several applications that are meant for basic use.
We compare them using several criteria to find out the strengths and weaknesses of the current implementations and identify a potential space for improvements.
Finally, we devote our attention to our own implementation.
We explain the proposed desktop software and go into details on software engineering aspects.
Specifically, modularity, architecture, usability, localization, testing, and automation.
In the end, we compare it with the existing software and identify possible future improvements.
Another result we focus on is a supporting website and how it helps with the use of the software.
All in a way open for extensibility by the community.

@@

There are several preconditions that allow us to consider electronic signatures as an alternative.
Firstly, the current state of the law and recent changes relating to the use of electronic signatures, secondly, well-established concepts in cryptography like \pojem{asymmetric cryptography} or \pojem{hashing}, and lastly, expertise from the field of software engineering - design, implementation, and maintenance of computer software.
We will concetrate on the first two in relation to the electronic signatures, and, where applicable, open-source desktop computer software.

@@

In general, electronic signature can be represented in different ways (e.g. as an image, text or other data attached to the document) and they \textbf{can not be denied in legal proceedings} just because they are in the electronic form \cite{eidas}.

\pojem{Advanced electronic signatures} are a subset of electronic signatures that have to \textbf{uniquely link and identify} the signature author, be created in a way that is \textbf{possible only by them} and any \textbf{changes to the signed document have to be detectable} \cite{eidas}.
From the technical point of view, standards \pojem{PAdES}, \pojem{XAdES}, and \pojem{CAdES} specified by the European Telecommunications Standards Institute (ETSI) comply with these requirements. We explain why this is the case in Section \ref{cryptography}.

\pojem{Qualified electronic signatures} (QES) are a subset of \pojem{advanced electronic signatures} with more specific requirements that can be found in annex 1 of the eIDAS regulation\footnote{Available at https://eur-lex.europa.eu/legal-content/EN/TXT/HTML/?uri=CELEX:32014R0910\#d1e32-111-1.}.
In more practical terms, such \pojem{advanced electronic signatures} are created with a qualified device using a qualified certificate and a qualified trust service.
Qualified, in this case, means authorized for such use by the legal authorities.
These devices, \pojem{certificates}, and trust services are explicitly listed by the EU. In many cases indirectly by listing the organization (and its items) from a specific country.

@@

Mentioned terminology is used when considering a natural person. Legal entities are able to use \pojem{electronic seals}, which are, from the technical point of view and for our purposes, identical.

@@

Such right can be exercised to license the software under either \pojem{proprietary} or \pojem{open-source} license.
\pojem{Proprietary} meaning under the exclusive legal right of the author, typically also \textbf{confidential and distributed as a paid product}.
\pojem{Open-source} meaning having its source code \textbf{freely available}, typically also \textbf{distributed for free} and without any liability (the software is provided "as is").

Motivation to license the software under an \pojem{open-source} license can differ, and so do such licenses.
While a \pojem{proprietary} license is usually made specifically for that entity and its interests, \pojem{open-source} licenses tend to be reused between different authors.
This means that consumers of the software can quickly recognize their rights and responsibilities if they decide to use, modify, or distribute the software.
In general, we can classify the \pojem{open-source} licenses into two categories: \pojem{permissive} and \pojem{copyleft}.

\pojem{Copyleft} license, in general, requires the user to publish the modified work under the compatible (free) license.
As a result, it forces them to extend the rights they have received onto others and, in turn, somehow hinder usability in software licensed under a different license.
A popular example of such license is the GNU General Public License(GPL) and its derivatives like GNU Lesser GPL (LGPL) or GNU Affero GPL (AGPL).

\pojem{Permissive} licenses, on the other hand, do not have such a requirement in place and are more suitable for potential commercialization.
They typically allow commercial use, modifications, distribution or sublicensing as long as the author is not held liable, and the original copyright and license are distributed with the software.
Often used permissive licenses are MIT License, Apache License, or BSD License.

@@

As we have already mentioned, software licenses can influence whether and how it can be used as a part of different software.
Both when we are the ones using someone else's software and when someone else is using our software.
Generally speaking, the more open the license is, the higher the probability it will not prevent adoption for legal reasons.
Various possible combinations can be seen in Figure \ref{fig:license-compatibility}, where greyed out options mean such combination is not possible unless a separate licensing agreement is reached with the copyright owner.

\obrazok{./figures/license-compatibility}{Schematic representation of license directionality. \osoba{Morin 2012} \cite{licensing}}{license-compatibility}

@@

\textbf{Secure distribution of keys} seems to be feasible when we are adding a new user to our network.
We require each user to personally come and pick up a physical device that contains pre-generated keys.
Changes made after that, however, would mean we either need to require a periodic personal visit to update the saved keys or transmission of these keys over some medium, which can be potentially unsafe.
Any new members of the network are unable to participate until other Alices and Bobs get their updates.

Even if we think we have established a reasonably secure way to do that and we are willing to wait, we realize the \textbf{number of keys} is quickly increasing with each new member. With \textit{n} users, the number of keys is

@@

\textbf{Either party can also lie} about not signing the document.
Since both have the key that can be used for encryption, they can sign any document on behalf of the other party, and there is no way to disprove it if we look at it as a third party.

These issues can be addressed with \pojem{asymmetric cryptography}.
In general, we assign \textbf{a pair of keys} to each person, one of them used for encryption, known as a \pojem{public key}, another used for decryption, known as a \pojem{private key}.
One party can encrypt, and the other is the only one that can decrypt.

\pojem{Digital signatures} are specific \pojem{asymmetric cryptography} algorithms.
The signer is the only one who possesses the \pojem{private key} used for signing, and others can use the \pojem{public key} for signature verification.
We \textbf{sign} the document, including the other parties' signature.

@@

Practical, widely used implementation of this idea is the \pojem{RSA} signature scheme that relies on the integer factorization problem (difficulty of factoring a product of two large primes) published in 1978 \cite{rsa}.

@@

We are left with the \textbf{problem of public key distribution} - even though \pojem{public-key schemes} do not require a \textit{secure channel}, they require \textit{authenticated channels} for the distribution of the public keys \cite{cryptotxtbook}.

A common solution is the use of \pojem{certificates}. \pojem{Certificates} are essentially \pojem{digital signatures} with metadata that can be used to establish the validity of the received \pojem{public key} before it is used.
Distribution of such \pojem{certificates} is the responsibility of a \pojem{Certification Authority} (CA) - a third party that all users in the network trust.
In the case of the Slovak ID cards, it is Disig - SVK eID Accredited CA, issued by the National Security Authority\footnote{Translation from "Národný bezpečnostný úrad".}.
Such \pojem{certificates}, therefore, \textbf{reliably uniquely link and identify} the \pojem{public key} owners, as can be seen in Figure \ref{fig:certificate-scheme}\footnote{Created by an unknown author, licensed under the CC BY-SA 3.0 license.}.
Commonly used standard for public key certificates is X.509\footnote{Specification available under RFC5280 at \url{https://tools.ietf.org/html/rfc5280}.}.

@@

It is often useful to map data of arbitrary size to a fixed-length (typically short) set of bits.
This process is commonly called \pojem{hashing}, a function that can be used for it \pojem{hash function}, and output of such function \pojem{hash}.

In cryptography, we are considering only \pojem{hash functions} that can also satisfy these additional \textbf{requirements}:

\begin{enumerate}
  \item \textbf{Deterministic} - same input data are always mapped to the same \pojem{hash}.
  \item \textbf{One-way} - it should be practically impossible to retrieve input data for a given \pojem{hash}.
  \item \textbf{Resistant to collisions} - it should be practically impossible to find two inputs with the same \pojem{hash}.
  \item Even small change to the input should lead to significantly different \pojem{hash} - so-called \textbf{avalanche effect}.
\end{enumerate}

In the context of \pojem{digital documents}, we can use such \pojem{hash function} to generate a \pojem{hash} for the document.
This \pojem{hash} can represent a document of any size and is used during the creation of a \pojem{digital signature} instead of the document itself.
Any further changes to the document will, therefore, lead to a significantly different \pojem{hash}, which means the \pojem{digital signature} will no longer be valid, and these \textbf{changes are detected}.
For this purpose, only \pojem{hash functions} that are reasonably fast can be used, so that we can utilize them on documents of variable size.

@@

An example of commonly used \pojem{hash function} family is \pojem{Secure Hash Algorithms} (SHA), with \pojem{hash functions} relevant to this thesis being mainly SHA-2\footnote{SHA-2 specification available in FIPS PUB 180-4 at \url{https://nvlpubs.nist.gov/nistpubs/FIPS/NIST.FIPS.180-4.pdf}.}.

@@

A common part of the paper signed documents is the date when they were signed.
Their digital counterpart is PKI-based \pojem{trusted timestamping} - a process of \pojem{digitally signing} date and time of document creation or modification.
No one should be able to tamper with such \pojem{timestamp} - not even the author of the document.
Additionally, we should not be able to choose arbitrary date and time ourselves, but it should represent the actual time at the signing.

To achieve this, a trusted third party - \pojem{Timestamping Authority} (TSA) - is the one that will sign the \pojem{hash} of the document together with the current date and time.
This process can be seen in Figure \ref{fig:trusted-timestamping}\footnote{Schema created by Bart Van den Bosch, licensed under the CC BY-SA 2.0 BE license.}.
Verifying can be done by decrypting this information using the \pojem{public key} and comparing the \pojem{hashes}. Any attempt at \textbf{tampering is therefore detected}.

@@

Extensions for PDF files, XML files, and CMS signed data for signing using \pojem{advanced electronic signatures} (as explained in Section \ref{eidas}), are \pojem{PAdES}, \pojem{XAdES}, and \pojem{CAdES}, respectively.
They are defined in the respective specifications by the ETSI in \cite{pades}, \cite{xades}, and \cite{cades}.

@@

The ability to keep signed documents valid for a prolonged period of time is covered in these standards under optional \pojem{Long-Term Validation} (LTV) that allows \textbf{archiving} of documents for many years, possibly decades.
If enabled, a trustworthy timestamp is required as validity is verified for the time the signature (and therefore timestamp) was created.
The expiration time is then limited by the validity of the timestamp certificate.

@@

\begin{enumerate}
  \item \textbf{Licensing} - is it freely available - both in terms of money and source code - best case is open-source; worst case is paid proprietary software.
  \item \textbf{Cross-platform availability} - is it available for users of all major desktop platforms - best case is availability on Windows, Mac OS X, and Linux; worst case is availability only on one platform.
  \item \textbf{Localization} - is it translated and does it display data like dates in local format - best case is availability at least in English and Slovak language; worst case is availability only in one language with no support for localization of dates, etc..
  \item \textbf{File support} - does it support different file formats and sizes - best case is support for all file formats with sensible size up to 50 MB; worst case is availability only for one file format or strict limits on the file size.
  \item \textbf{Easy setup} - does it require non-trivial setup to use the SR eID - best case is no setup required; worst case is possibly demanding setup or troubleshooting.
  \item \textbf{Flexibility} - how much logic can be customized - best case is various options for signature format, signature policy, used algorithms or timestamping configuration; worst case is none or almost none customization.
  \item \textbf{Built-in verification} - is the verification of signatures available in the same software - best case is ability to verify various signature types; worst case is no support for signature verification.
  \item \textbf{Extensibility} - is it extensible by the community - best case is it is prepared for easy extensibility in terms of functionality and localization; worst case is no extensibility.
\end{enumerate}

@@

The application consists only of one window, which makes it very simple at first sight.
The very first input - "Keystore type" - can be harder to understand for the basic user since it contains over a dozen options in the form of abbreviations like "CASEEXACTJKS" or "PKCS12-DEF-3DES-40RC2".
The default value is "WINDOWS-MY", however, which is what is exactly required to use the SR eID.
Application has a toggle for the "Advanced view" that is off by default, but it is necessary to turn it on if we want to choose the correct key.
On some of the computers, we were not able to see the key from the eID. On others, it took a while and several pressing of a button to load them.
The correct key was not the first, preselected key in the form of "Name Surname", but the second "Name Surname (1)".
With everything set up correctly, the application was successfully able to attach a QES.
Even though the application looks pretty simple, it is still pretty flexible with its options regarding the used algorithms, TSA, OCSP, or CRL.
That being said, it is available only in English, and we have also noticed issues with the diacritics that you can see in Figure \ref{fig:screenshot-jsignpdf} - "Jakub Ďuraš" is displayed as "Jakub Ïura".

@@

\obrazok{./figures/screenshot-adobeacrobat}{Screenshot of the Adobe Acrobat Reader DC}{screenshot-adobeacrobat}

Using the QES is slightly harder to find in the Adobe Acrobat Reader.
It is not available under "Signing", but "More Tools" where it is called "Certificates".
Setup of the certificate for use with the SR eID can be a hit or miss.
On one computer, the application would not include a certificate from the SR eID, and we had to go through a process of adding the PKCS \#11 library ourselves.
That meant we had to go through several steps and windows and disable "protected mode".
The Setup of this application should be done for the user ideally by the IT personnel.
On another computer, we were simply able to choose from the list of the available certificates.
That being said, everything is correctly localized, and being one of the most used applications of its kind, it is very easy to search for help in situations like this.
Outside of that, verification of signatures is part of the application and is very responsive and flexible.
The application is proprietary and can not be extended or verified.
It is also available only on the Windows and Mac OS X.

@@

We have summarized our findings in Table \ref{table:software}. Value \textit{X} means best-case scenario, no value worst-case scenario, and value \textit{/} somewhere in between - partially or depends.

\begin{table}[h!]
  \begin{tabular}{ p{3cm}|p{1cm}p{1cm}p{1cm}p{1cm}p{1cm}p{1cm}p{1cm}p{1cm} }
                      & \rot{licensing} & \rot{cross-platform} & \rot{localization} & \rot{file suppport} & \rot{easy setup} & \rot{flexibility} & \rot{built-in verification} & \rot{extensibility} \\
    \hline
    JSignPdf          & X               & /                   &                    &                      &                  & X                 &                             & / \\
    zep.disig.sk      & /               & X                   & X                  & /                    & X                &                   & X                           &   \\
    Acrobat DC        & /               & /                   & X                  &                      & /                & X                 & X                           &   \\
    D.PDF Signer      &                 &                     & /                  &                      & X                & /                 &                             &   \\
  \end{tabular}

@@

@@

The main module is an Electron\footnote{Electron is a framework allowing the development of desktop software using web technologies.} application written in the TypeScript.
This module features simple UI and abstraction around modules for signing that we call \pojem{backends}.
It handles all of the communication with the user for the backend and tries to simplify development of the backends as much as possible.

Signing modules - \pojem{backends} - are separate CLI applications written in any programming language.
They are responsible for the actual document manipulation - signing and verification.

@@

There are several advantages to this approach.
Firstly, we can write backends in any language that suits us the most.
For example, our backend that allows signing documents with the signatures compliant with the EU eIDAS directive is written in Java because it is based on a library from EU - DSS - written in Java.
We expect most of the languages have APIs for the consumption of the standard streams with excellent portability across different desktop platforms.
Secondly, since we are working on separate applications, they can be written by different people under a different license.
That also means we can easily swap the main module or any of the backends for use in different environments or different bundles.
Thirdly, together with the choice of using standard streams, it means that anyone can quickly start the development of a new backend.
Writing to the standard output stream in CLI application is one of the first things developers usually learn - so-called "Hello World" applications.
It usually should not involve studying any required third-party libraries, and authors can choose the language they are most familiar with.

@@

\subsection{Usability}

\subsection{Localization}

To make our application accessible to as many users as possible, we have also made the application translatable and try to display data like dates in the local format.

>>>>

\usepackage{amssymb}
\newcommand\doublecheck{\checkmark\kern-0.6em\checkmark}

@@

\abstrakt{
So zmenami v právnom statuse elektronických podpisov vo svete sa vytvára príležitosť rozšíriť elektronické podpisovanie ako plnohodnotnú náhradu pre vlastnoručné podpisy.
Elektronické podpisy sú dôležité čoraz viac v dobe elektronickej komunikácie, kedy sú vlastnoručné podpisy nepraktické z dôvodu nutnosti tlače, osobného kontaktu alebo zasielania podpísaných dokumentov.
V práci sa venujeme právnym predpokladom relevantným pre EÚ, technologickým predpokladom, prehľadu v súčasnosti dostupného softvéru a návrhu vlastnej softvérovej platformy.
Výsledkom je otvorený, používateľsky prívetivý softvér pre platformy Windows, Mac OS X a Linux.
Pomocou aplikácie je možné, v súlade s nariadením eIDAS, podpísať a overiť podpis ľubovoľného dokumentu, ktorý má rovnaký právny účinok ako osvedčený podpis.
Naše riešenie podporuje občianske preukazy Slovenskej Republiky s čipom eID bez akýchkoľvek úprav alebo nastavení, čím sa umožní ich jednoduché využitie mimo obvyklého použitia vo webových aplikáciách štátu.
Aplikácia nie je viazaná iba na slovenské eID, ale je pripravená na ďalšie použitie s rôznymi zariadeniami v rôznych štátoch a ako platforma pre iné druhy podpisov.
Súčasťou je aj webová stránka, ktorá okrem možnosti stiahnutia softvéru a stručnej dokumentácie poskytuje všeobecné informácie o elektronických podpisoch, aby bol projekt čo najviac prístupný verejnosti.
Dôraz kladieme na jednoduchú rozšíriteľnosť programu vo forme jazykových prekladov, dokumentácie, funkcionality a všeobecných informácií.
}

@@

While, specifically in the European Union, electronic signatures seem to be relatively extensively used in communication with the government, they do not seem to be widely adopted in the exchange between individuals or smaller organizations.
The problem is that handwritten signatures are often unpractical for electronic communication due to the need of physical presence or printing and sending of documents.

Our goal is to explore an accessible way to use electronic signatures.
In the beginning, we go over the preconditions that make it a viable alternative to the handwritten signatures.
We cover the legal preconditions relevant for the European Union and Slovakia in particular.
Also, we explain technical preconditions focusing on the cryptographic concepts like asymmetric cryptography, timestamping, hashing, and standards that comply with the law within the European Union.
Secondly, we do a review of the existing software, where we briefly cover several applications that are meant for basic use.
We compare them using several criteria to find out the strengths and weaknesses of the current implementations and identify a potential gap.
Finally, we devote our attention to our own implementation.
We explain the proposed desktop software and go into details on software engineering aspects.
Specifically, modularity, architecture, usability, localization, testing, and automation.
In the end, we compare it with the existing software and identify possible future improvements.
Another result we focus on is a supporting website and how it helps with the use of the software.
All in a way that is open for extensibility by the community.

@@

There are several preconditions that allow us to consider electronic signatures as an alternative.
Firstly, the current state of the law and recent changes relating to the use of electronic signatures, secondly, well-established concepts in cryptography like \pojem{asymmetric cryptography} or \pojem{hashing}, and lastly, expertise from the field of software engineering - design, implementation, and maintenance of computer software.
We will concentrate on the first two in relation to the electronic signatures and where applicable open-source desktop computer software.

@@

In general, electronic signature can be represented in different ways (e.g. as an image, text or other data attached to the document) and they can not be denied in legal proceedings just because they are in the electronic form \cite{eidas}.

\pojem{Advanced electronic signatures} are a subset of electronic signatures that have to uniquely link and identify the signature author, be created in a way that is possible only by them and any changes to the signed document have to be detectable \cite{eidas}.
From the technical point of view, standards \pojem{PAdES}, \pojem{XAdES}, and \pojem{CAdES} specified by the European Telecommunications Standards Institute (ETSI) comply with these requirements. We explain why this is the case in Section \ref{cryptography}.

\pojem{Qualified electronic signatures} (QES) are a subset of \pojem{advanced electronic signatures} with more specific requirements that can be found in annex 1 of the eIDAS regulation\footnote{Available at https://eur-lex.europa.eu/legal-content/EN/TXT/HTML/?uri=CELEX:32014R0910\#d1e32-111-1.}.
In more practical terms, such advanced electronic signatures are created with a qualified device using a qualified certificate and a qualified trust service.
Qualified, in this case, means authorized for such use by the legal authorities.
These devices, certificates, and trust services are explicitly listed by the EU. In many cases indirectly by listing the organization (and its items) from a specific country.

@@

Mentioned terminology is used when considering a natural person. Legal entities are able to use \pojem{electronic seals}, which are, from the technical point of view and for our purposes, identical to electronic signatures.

@@

Such right can be exercised to license the software under either \pojem{proprietary} or \pojem{open-source} license.
\iffalse % Not part of the version for ŠVK
\pojem{Proprietary} meaning under the exclusive legal right of the author, typically also confidential and distributed as a paid product.
\pojem{Open-source} meaning having its source code freely available, typically also distributed for free and without any liability (the software is provided "as is").

Motivation to license the software under an \pojem{open-source} license can differ, and so do such licenses.
While a \pojem{proprietary} license is usually made specifically for that entity and its interests, \pojem{open-source} licenses tend to be reused between different authors.
This means that consumers of the software can quickly recognize their rights and responsibilities if they decide to use, modify, or distribute the software.
\fi
In general, we can classify the \pojem{open-source} licenses into two categories: \pojem{permissive} and \pojem{copyleft}.

\pojem{Copyleft} license, in general, requires the user to publish the modified work under the compatible (free) license.
As a result, it forces them to extend the rights they have received onto others and, in turn, somehow hinder usability in software licensed under a different license.
\iffalse % Not part of the version for ŠVK
A popular example of such license is the GNU General Public License(GPL) and its derivatives like GNU Lesser GPL (LGPL) or GNU Affero GPL (AGPL).
\fi

\pojem{Permissive} licenses, on the other hand, do not have such a requirement in place and are more suitable for potential commercialization.
They typically allow commercial use, modifications, distribution or sublicensing as long as the author is not held liable, and the original copyright and license are distributed with the software.
\iffalse % Not part of the version for ŠVK
Often used permissive licenses are MIT License, Apache License, or BSD License.
\fi

@@

\iffalse % Not part of the version for ŠVK
As we have already mentioned, software licenses can influence whether and how it can be used as a part of different software.
Both when we are the ones using someone else's software and when someone else is using our software.
\fi
Generally speaking, the more open the license is, the higher the probability it will not prevent adoption for legal reasons.
We provide examples of categorized licenses in Figure \ref{fig:license-comparison}.

\iffalse % Not part of the version for ŠVK
Various possible combinations can be seen in Figure \ref{fig:license-compatibility}, where greyed out options mean such combination is not possible unless a separate licensing agreement is reached with the copyright owner.

\obrazok{./figures/license-compatibility}{Schematic representation of license directionality. \osoba{Morin 2012} \cite{licensing}}{license-compatibility}
\fi

@@

Secure distribution of keys seems to be feasible when we are adding a new user to our network.
We require each user to personally come and pick up a physical device that contains pre-generated keys.
Changes made after that, however, would mean we either need to require a periodic personal visit to update the saved keys or transmission of these keys over some medium, which can be potentially unsafe.
Any new members of the network are unable to participate until other Alices and Bobs get their updates.

Even if we think we have established a reasonably secure way to do that and we are willing to wait, we realize the number of keys is quickly increasing with each new member. With \textit{n} users, the number of keys is

@@

Either party can also lie about not signing the document.
Since both have the key that can be used for encryption, they can sign any document on behalf of the other party, and there is no way to disprove it if we look at it as a third party.

These issues can be addressed with \pojem{asymmetric cryptography}.
In general, we assign a pair of keys to each person, one of them used for encryption, known as a \pojem{public key}, another used for decryption, known as a \pojem{private key}.
One party can encrypt, and the other is the only one that can decrypt.

\pojem{Digital signatures} are specific \pojem{asymmetric cryptography} algorithms.
The signer is the only one who possesses the \pojem{private key} used for signing, and others can use the \pojem{public key} for signature verification.
We sign the document, including the other parties' signature.

@@

Practical, widely used implementation of this idea is the \pojem{RSA} signature scheme that relies on the integer factorization problem published in 1978 \cite{rsa}.

@@

We are left with the problem of public key distribution - even though \pojem{public-key schemes} do not require a \textit{secure channel}, they require \textit{authenticated channels} for the distribution of the public keys \cite{cryptotxtbook}.

A common solution is the use of \pojem{certificates}. \pojem{Certificates} are essentially \pojem{digital signatures} with metadata that can be used to establish the validity of the received \pojem{public key} before it is used.
Distribution of such \pojem{certificates} is the responsibility of a \pojem{Certification Authority} (CA) - a third party that all users in the network trust.
In the case of the Slovak ID cards, it is Disig - SVK eID Accredited CA, issued by the National Security Authority\footnote{Translation from "Národný bezpečnostný úrad".}.
Such \pojem{certificates}, therefore, reliably uniquely link and identify the \pojem{public key} owners, as can be seen in Figure \ref{fig:certificate-scheme}\footnote{Created by an unknown author, licensed under the CC BY-SA 3.0 license.}.
Ability to do that is the first requirement for the advanced electronic signatures mentioned in Section \ref{eidas}.
Commonly used standard for public key certificates is X.509\footnote{Specification available under RFC5280 at \url{https://tools.ietf.org/html/rfc5280}.}.

@@

It is often useful to map data of arbitrary size to a fixed-length and typically short set of bits called \pojem{hash}.
Function that can be used for it is called \pojem{hash function}.

In cryptography, we are considering only \pojem{hash functions} that can also satisfy these additional requirements:

\begin{enumerate}
  \item \textbf{Deterministic} - same input data are always mapped to the same \pojem{hash}.
  \item \textbf{One-way} - it should be infeasible to retrieve input data for a given \pojem{hash}.
  \item \textbf{Resistant to collisions} - it should be infeasible to find two inputs with the same \pojem{hash}.
  \item \textbf{Avalanche effect} - even small change to the input should lead to significantly different \pojem{hash}.
\end{enumerate}

In the context of \pojem{digital documents}, we can use such hash function to generate a hash for the document.
This hash can represent a document of any size and is used during the creation of a digital signature instead of the document itself.
Any further changes to the document will, therefore, lead to a significantly different hash, which means the digital signature will no longer be valid, and these changes are detected.
This ability is one of the requirements for the advanced electronic signatures from Section \ref{eidas}.
For this purpose, only hash functions that are reasonably fast can be used, so that we can utilize them on documents of variable size.

@@

An example of commonly used hash function family is \pojem{Secure Hash Algorithms} (SHA), with hash functions relevant to this thesis being mainly SHA-2\footnote{SHA-2 specification available in FIPS PUB 180-4 at \url{https://nvlpubs.nist.gov/nistpubs/FIPS/NIST.FIPS.180-4.pdf}.}.

@@

A common part of the paper signed documents is the date when they were signed.
Their digital counterpart is PKI-based \pojem{trusted timestamping} - a process of digitally signing date and time of document creation or modification.
No one should be able to tamper with such timestamp - not even the author of the document.
Additionally, we should not be able to choose arbitrary date and time ourselves, but it should represent the actual time at the signing.

To achieve this, a trusted third party - \pojem{Timestamping Authority} (TSA) - is the one that will sign the \pojem{hash} of the document together with the current date and time.
This process can be seen in Figure \ref{fig:trusted-timestamping}\footnote{Schema created by Bart Van den Bosch, licensed under the CC BY-SA 2.0 BE license.}.
Verifying can be done by decrypting this information using the public key and comparing the hashes.
Any attempt at tampering is therefore detected, which helps to cover requirements for the advanced electronic signatures from Section \ref{eidas}.

@@

Extensions for PDF files, XML files, and CMS signed data for signing using advanced electronic signatures, as explained in Section \ref{eidas}, are \pojem{PAdES}, \pojem{XAdES}, and \pojem{CAdES}, respectively.
They are defined in the respective specifications by the ETSI in \cite{pades}, \cite{xades}, and \cite{cades}.

@@

The ability to keep signed documents valid for a prolonged period of time is covered in these standards under optional \pojem{Long-Term Validation} (LTV) that allows archiving of documents for many years, possibly decades.
If enabled, a trustworthy timestamp is required as validity is verified for the time the signature (and therefore timestamp) was created.
The expiration time is then limited by the validity of the timestamp certificate.

@@

\begin{enumerate}
  \item \textbf{Licensing} - is it freely available - both in terms of money and source code - best case is open-source; worst case is paid proprietary software.
  \item \textbf{Cross-platform compatibility} - is it available for users of all major desktop platforms - best case is availability on Windows, Mac OS X, and Linux; worst case is availability only on one platform.
  \item \textbf{Localization} - is it translated and does it display data like dates in local format - best case is availability at least in English and Slovak language; worst case is availability only in one language with no support for localization of dates, etc..
  \item \textbf{File support} - does it support different file formats and sizes - best case is support for all file formats with sensible size up to 50 MB; worst case is availability only for one file format or strict limits on the file size.
  \item \textbf{Easy setup} - does it require non-trivial setup to use the SR eID - best case is no setup required; worst case is possibly demanding setup or troubleshooting.
  \item \textbf{Flexibility} - how much logic can be customized - best case is various options for signature format, signature policy, used algorithms or timestamping configuration; worst case is none or almost none customization.
  \item \textbf{Built-in verification} - is the verification of signatures available in the same software - best case is ability to verify various signature types; worst case is no support for signature verification.
  \item \textbf{Extensibility} - is it extensible by the community - best case is it is prepared for easy extensibility in terms of functionality and localization; worst case is no extensibility.
\end{enumerate}

@@

The application consists only of one window, which makes it very simple at first sight.
The very first input - \textit{Keystore type} - can be harder to understand for the basic user since it contains over a dozen options in the form of abbreviations like \textit{CASEEXACTJKS} or \textit{PKCS12-DEF-3DES-40RC2}.
The default value is \textit{WINDOWS-MY}, however, which is what is exactly required for the SR eID.
Application has a toggle for the \textit{Advanced view}, which is off by default, but it is necessary to turn it on if we want to choose the correct key.
On some of the computers, we were not able to see the key from the eID. On others, it took a while and several pressing of a button to load them.
The correct key was not the first, preselected key in the form of \textit{Name Surname}, but the second \textit{Name Surname (1)}.
With everything set up correctly, the application was successfully able to attach a QES.
Even though the application looks pretty simple, it is still pretty flexible with its options regarding the used algorithms, TSA, OCSP, or CRL.
That being said, it is available only in English, and we have also noticed issues with the diacritics as you can see in Figure \ref{fig:screenshot-jsignpdf} - "Jakub Ďuraš" is displayed as "Jakub Ïura".

@@

\obrazok{./figures/screenshot-adobeacrobat}{Screenshot of the Adobe Acrobat Reader DC.}{screenshot-adobeacrobat}

Using the QES is slightly harder to find in the Adobe Acrobat Reader.
It is not available under \textit{Signing}, but \textit{More Tools} where it is called \textit{Certificates}.
Setup of the certificate for use with the SR eID can be a hit or miss.
On one computer, the application would not include a certificate from the SR eID, and we had to go through a process of adding the PKCS \#11 library ourselves.
That meant we had to go through several steps and windows and disable the \textit{protected mode} option.
The setup of this application should be done for the user ideally by the IT personnel.
On another computer, we were simply able to choose from the list of the available certificates.
That being said, everything is correctly localized, and being one of the most used applications of its kind, it is very easy to search for help in situations like this.
Outside of that, verification of signatures is part of the application and is very responsive and flexible.
The application is proprietary and can not be extended or verified.
It is available only on the Windows and Mac OS X.

@@

We have summarized our findings in Table \ref{table:software}. Value \doublecheck means best-case scenario, no value worst-case scenario, and value \checkmark somewhere in between - partially or depends.

\begin{table}[h!]
  \begin{tabular}{ p{2.75cm}|p{0.85cm}p{0.85cm}p{0.85cm}p{0.85cm}p{0.85cm}p{0.85cm}p{0.85cm}p{0.85cm} }
                      & \rot{licensing} & \rot{cross-platform} & \rot{localization} & \rot{file suppport} & \rot{easy setup} & \rot{flexibility} & \rot{built-in verification} & \rot{extensibility} \\
    \hline
    JSignPdf          & \doublecheck    & \checkmark           &                    &                     &                  & \doublecheck      &                             & \checkmark \\
    zep.disig.sk      & \checkmark      & \doublecheck         & \doublecheck       & \checkmark          & \doublecheck     &                   & \doublecheck                &            \\
    Acrobat DC        & \checkmark      & \checkmark           & \doublecheck       &                     & \checkmark       & \doublecheck      & \doublecheck                &            \\
    D.PDF Signer      &                 &                      & \checkmark         &                     & \doublecheck     & \checkmark        &                             &            \\
  \end{tabular}

@@

\subsection{Usability and localization}

We have also made the application easily translatable and try to display data like dates in the local format.
When the application is started, we try to detect the locale of the user and set it as a default language.
If we do not yet support the locale of the user, we default to the English language.
At the time of writing, our application is fully translated in three languages: English (source language), Slovak, and Czech.

@@

The main module is an Electron\footnote{Electron is a framework allowing the development of desktop software using web technologies.} application written in the language TypeScript\footnote{TypeScript is statically typed superset of JavaScript that compiles to JavaScript.}.
This module features simple UI and abstraction around modules for signing that we call \pojem{backends}.
It handles all of the communication with the user for the backend and tries to simplify development of the backends as much as possible.

Signing modules - backends - are separate CLI applications written in any programming language.
They are responsible for the actual document manipulation - signing and verification.

@@

There are several advantages to this approach.
Firstly, we can write backends in any language that suits us the most.
For example, our backend that allows signing documents with the signatures compliant with the EU eIDAS directive is written in Java because it is based on a library from EU - DSS - also written in Java.
We expect most of the languages have APIs for the consumption of the standard streams with excellent portability across different desktop platforms.
Secondly, since we are working on separate applications, they can be written by different people under a different license.
That also means we can easily swap the main module or any of the backends for use in different environments or different bundles.
Thirdly, together with the choice of using standard streams, it means that anyone can quickly start the development of a new backend.
Writing to the standard output stream in CLI application is one of the first things developers usually learn - so-called "Hello World" applications.
It usually should not involve studying any required third-party libraries, and authors can choose the language they are most familiar with.

@@
